---
title: "Methods 2 -- Portfolio Assignment 1"
output:
  pdf_document: default
---

- _Type:_ Group assignment
- _Due:_ 10 March 2024, 23:59

---

In the following exercises, you will be asked to generate and summarize simulations from statistical models. You should use what you have learned so far (i.e. for loop, if else statements, sampling from continuous and discrete distributions...) to generate observations and summarize your samples using (one of) the appropriate methods. You can find examples of how to do that in Ch. 5. Note that here we will only focus on generative models, several aspects for inference and hypothesis testing discussed in Ch. 4 are not directly needed in this context.

In the first exercise, we will assume that the population of interest has a proportion of 0.51 men and 0.49 women. Your model should reflect that.

Please submit your answers on GitHub Classroom.

---

1. _(5.2 from ROS)_ __Continuous probability simulation:__ The logarithms of weights (in pounds) of men in the United States are approximately normally distributed with mean 5.13 and standard deviation 0.17; women’s log weights are approximately normally distributed with mean 4.96 and standard deviation 0.20. Suppose 10 adults selected at random step on an elevator with a capacity of 1750 pounds. What is the probability that their total weight exceeds this limit?

```{r}

# Set seed:
set.seed(666)

# Define the means and sd's for weight (in pounds) of men and women. As they are logarithms use the exp():
m_mean <- exp(5.13)
m_sd <- exp(0.17)
w_mean <- exp(4.96)
w_sd <- exp(0.20)

n_sims <- 10 # number of simulations
elev_capacity <- 1750 # elevator capacity
exceed_capacity <- rep(NA, n_sims) # counter for the number of times the capacity is exceeded





```

---

2. _(5.6 from ROS)_ __Propagation of uncertainty:__ We use a highly idealized setting to illustrate the use of simulations in combining uncertainties. Suppose a company changes its technology for widget production, and a study estimates the cost savings at \$5 per unit, but with a standard error of \$4. Furthermore, a forecast estimates the size of the market (that is, the number of widgets that will be sold) at 40 000, with a standard error of 10 000. Assuming these two sources of uncertainty are independent, use simulation to estimate the total amount of money saved by the new product (that is, savings per unit, multiplied by size of the market).

```{r}

mean_savings <- 5 # mean cost savings per unit
se_savings <- 4 # standard error of cost savings per unit
market_size <- 40000 # mean market size
market_se <- 10000 # standard error of market size

# create simulation

# 

```


---

3. _(5.10 from ROS)_ __Inference for a ratio of parameters:__ A (hypothetical) study compares the costs and effectiveness of two different medical treatments.

    - In the first part of the study, the difference in costs between treatments A and B is estimated at \$600 per patient, with a standard error of \$400, based on a regression with 50 degrees of freedom.
    - In the second part of the study, the difference in effectiveness is estimated at 3.0 (on some relevant measure), with a standard error of 1.0, based on a regression with 100 degrees of freedom.
    - For simplicity, assume that the data from the two parts of the study were collected independently.

    Inference is desired for the incremental cost-effectiveness ratio: the difference between the average costs of the two treatments, divided by the difference between their average effectiveness, a problem discussed further by Heitjan, Moskowitz, and Whang (1999).

    (a) Create 1000 simulation draws of the cost difference and the effectiveness difference, and make a scatterplot of these draws.
    (b) Use simulation to come up with an estimate, 50% interval, and 95% interval for the incremental cost-effectiveness ratio.
    (c) Repeat, changing the standard error on the difference in effectiveness to 2.0.

```{r 5.10 a - simulation and scatterplot}

library(pacman)
p_load(ggplot2,
       tidyverse)

# set seed
set.seed(420)

# Generate t-distribution for the cost:
n <-  1000
df_cost <- 50
t_dist_cost <- rt(n = n, df = df_cost)

# Define desired mean and se for cost:
desired_mean_cost <- 600
desired_se_cost <- 400 # Should we convert se to sd before proceeding here???

# Scale the t-distribution to the desired mean and se:
t_dist_cost_scale <- scale(t_dist_cost)*desired_se_cost + desired_mean_cost # the scale() function "resets" sd = 1 and mean = 0, so now we can use our own desired values instead

# Generate t-distribution for the effectiveness:
n <-  1000
df_eff <- 100
t_dist_eff <- rt(n = n, df = df_eff)

# Define desired mean and se for effectiveness:
desired_mean_eff <- 3.0
desired_se_eff <- 1.0

# Scale the t-distribution to the desired mean and se:
t_dist_eff_scale <- scale(t_dist_eff)*desired_se_eff + desired_mean_eff

# Create a dataframe to make the scatter plot:
data_cost_eff_diff <- data.frame(
  cost_diff = t_dist_cost_scale,
  eff_diff = t_dist_eff_scale
  )

# Create new columns for "cheapest" and "most effective":
data_cost_eff_diff <- data_cost_eff_diff %>% mutate(
  cheapest = ifelse(cost_diff > 0, "A", "B"),
  most_effective = ifelse(eff_diff < 0, "A", "B"))

# Create the scatterplot:
plot_t <- ggplot(data_cost_eff_diff, aes(
  x = cost_diff,
  y = eff_diff,
  color = cheapest,
  shape = most_effective
  )) +
  geom_point() +
  labs(
    title = "Scatterplot of cost and effectiveness differences",
    x = "Cost difference",
    y = "Effectiveness difference",
    color = "Cheapest treatment",
    shape = "Most effective treatment"
  ) + 
  theme_minimal()

# Show plot:
plot_t

```

```{r 5.10 b - get ratio, 50% and 95% intervals}

# Calculate the incremental cost-effectiveness ratio:
data_cost_eff_diff <- data_cost_eff_diff %>% 
  mutate(
  ice_ratio = cost_diff / eff_diff
  )

# Get the mean and sd of the ice_ratio:
ice_ratio_mean <- mean(data_cost_eff_diff$ice_ratio)
ice_ratio_sd <- sd(data_cost_eff_diff$ice_ratio)

# Convert sd to se:
ice_ratio_se <- ice_ratio_sd / sqrt(1000)

# Calculate the 50% interval:
ice_ratio_50 <- quantile(data_cost_eff_diff$ice_ratio, probs = c(0.25, 0.75))

# Calculate the 95% interval:
ice_ratio_95 <- quantile(data_cost_eff_diff$ice_ratio, probs = c(0.025, 0.975))

# Print the mean, sd and se:
print(paste("mean:", ice_ratio_mean, "sd:", ice_ratio_sd, "se:", ice_ratio_se))
print(paste("50% int quantile:", ice_ratio_50, "95% int quantile:", ice_ratio_95))

```

```{r 5.10 c - repeat with new se}

# Set seed:
set.seed(420)

# Generate t-distribution for the cost:
n <-  1000
df_cost_new <- 50
t_dist_cost_new <- rt(n = n, df = df_cost_new)

# Define desired mean and se for cost:
desired_mean_cost_new <- 600
desired_se_cost_new <- 400 # Should we convert se to sd before proceeding here???

# Scale the t-distribution to the desired mean and se:
t_dist_cost_scale_new <- 
  scale(t_dist_cost_new)*desired_se_cost_new + desired_mean_cost_new # the scale() function "resets" sd = 1 and mean = 0, so now we can use our own desired values instead

# Generate t-distribution for the effectiveness:
n <-  1000
df_eff_new <- 100
t_dist_eff_new <- rt(n = n, df = df_eff_new)

# Define desired mean and se for effectiveness:
desired_mean_eff_new <- 3.0
desired_se_eff_new <- 2.0

# Scale the t-distribution to the desired mean and se:
t_dist_eff_scale_new <- 
  scale(t_dist_eff_new)*desired_se_eff_new + desired_mean_eff_new

# Create a data frame to make the scatter plot:
data_cost_eff_diff_new <- data.frame(
  cost_diff = t_dist_cost_scale_new,
  eff_diff = t_dist_eff_scale_new
  )

# Create new columns for "cheapest" and "most effective":
data_cost_eff_diff_new <- data_cost_eff_diff_new %>% mutate(
  cheapest = ifelse(cost_diff > 0, "A", "B"),
  most_effective = ifelse(eff_diff < 0, "A", "B"))

# Create the scatter plot:
plot_new <- ggplot(data_cost_eff_diff_new, aes(
  x = cost_diff,
  y = eff_diff,
  color = cheapest,
  shape = most_effective)) +
  geom_point() +
  labs(
    title = "New scatterplot of cost and effectiveness differences",
    x = "Cost difference",
    y = "Effectiveness difference",
    color = "Cheapest treatment",
    shape = "Most effective treatment") + 
  theme_minimal()

# Show plot:
plot_new

# Calculate the incremental cost-effectiveness ratio:
data_cost_eff_diff_new <- data_cost_eff_diff_new %>% mutate(
  ice_ratio = cost_diff / eff_diff
  )

# Get the mean and sd of the ice_ratio:
ice_ratio_mean_new <- mean(data_cost_eff_diff_new$ice_ratio)
ice_ratio_sd_new <- sd(data_cost_eff_diff_new$ice_ratio)

# Convert sd to se:
ice_ratio_se_new <- ice_ratio_sd_new / sqrt(1000)

# Calculate the 50% interval:
ice_ratio_50_new <- quantile(data_cost_eff_diff_new$ice_ratio, probs = c(0.25, 0.75))

# Calculate the 95% interval:
ice_ratio_95_new <- quantile(data_cost_eff_diff_new$ice_ratio, probs = c(0.025, 0.975))

# Print the mean, sd and se:
print(paste("mean:", ice_ratio_mean_new, "sd:", ice_ratio_sd_new, "se:", ice_ratio_se_new))
print(paste("50% int quantile:", ice_ratio_50_new, "95% int quantile:", ice_ratio_95_new))

```

---
